Transfer of Value Functions via Variational Methods

### 2018-05-18 NIPS 2018/main.tex
delta

\begin{abstract}
We consider the problem of transferring value functions in reinforcement learning. We propose an approach that uses the given source tasks to learn a prior distribution over optimal value functions and provide an efficient variational approximation of the corresponding posterior in a new target task. We show our approach to be general, in the sense that it can be combined with complex parametric function approximators and distribution models, while providing two practical algorithms based on Gaussians and Gaussian mixtures. We theoretically analyze them by deriving a finite sample analysis and provide a comprehensive empirical evaluation in four different domains.
\end{abstract}

@@

\begin{restatable}{theorem}{thmg} \label{th:main-mixture}
Fix a target task $\mathcal{M}_{\tau}$ and let $\wt{Q}$ be the fixed-point of the corresponding mellow Bellman operator. Assume linearly parameterized value functions $Q_{\bm{w}}(s,a)=\bm{w}^T\bm{\phi}(s,a)$ with bounded weights $\norm{\bm{w}}_2\leq w_{max}$ and uniformly bounded features $\norm{\bm{\phi}(s,a)}_2\leq \phi_{\max}$. Consider the mixture version of Algorithm~ \ref{alg:transfer} using $C$ components, source task weights $\mathcal{W}_s$, and bandwidth $\sigma^2_p\ $ for the prior. Denote by $\wh{\bm{\xi}} = (\wh{\bm{\mu}}_1,\dots,\wh{\bm{\mu}}_C,\wh{\bm{\Sigma}}_1,\dots,\wh{\bm{\Sigma}}_C)$ the variational parameters minimizing the objective of Eq. \eqref{eq:elbo} on a dataset $D$ of $N$ i.i.d. samples distributed according to $\tau$ and $\nu$. Let $\bm{w}^* = \arginf_{\bm{w}}||\wt{B}_{\bm{w}}||_{\nu}^2$ and define $\upsilon(\bm{w}^*) \triangleq \mathbb{E}_{\mathcal{N}(\bm{w}^*,\frac{1}{N}\bm{I})}\left[ \text{v}(\bm{w}) \right]$, with $\text{v}(\bm{w}) \triangleq \mathbb{E}_{\nu}\left[Var_{\mathcal{P}_{\tau}}\left[\wt b(\bm{w})\right]\right]$. Then, there exist constants $c_1,c_2,c_3$ such that, with probability at least $1-\delta$ over the choice of weights $\bm{w} \sim \frac{1}{C}\sum_i\mathcal{N}(\wh{\bm{\mu}}_i,\wh{\bm{\Sigma}}_i)$ and dataset $D$:
\begin{equation}
\resizebox{0.93\textwidth}{!}{$
\mathbb{E}_{q_{\wh{\bm{\xi}}}}\left[ \norm{\wt B_{\bm{w}}}_{\nu}^2 \right ] \leq\  2\norm{\wt{B}_{\bm{w}^{*}}}_{\nu}^2 + \upsilon(\bm{w}^*) + c_1\sqrt{\frac{\log\frac{2}{\delta}}{N}} + \frac{c_2 + \lambda d \log N + 2\lambda\varphi\left(\frac{1}{2\sigma^2_p}\norm{\bm{w}^* - \bm{w}_j}\right)}{N} + \frac{c_3}{N^2}$},
\end{equation}
where, for a vector $\bm{x} = (x_1,\dots,x_d)$, $\varphi \left( x_j\right) \triangleq \sum_i \frac{e^{-x_i}}{\sum_j e^{-x_j}}x_i$ is the \textit{softmin} function.
\end{restatable}

@@

\begin{restatable}{theorem}{thg} \label{th:main-gaussian}
Fix a target task $\mathcal{M}_{\tau}$ and let $\wt{Q}$ be the fixed-point of the corresponding mellow Bellman operator. Assume linearly parameterized value functions $Q_{\bm{w}}(s,a)=\bm{w}^T\bm{\phi}(s,a)$ with bounded weights $\norm{\bm{w}}_{2}\leq w_{max}$ and uniformly bounded features $\norm{\bm{\phi}(s,a)}_{2}\leq \phi_{\max}$. Consider the Gaussian version of Algorithm~\ref{alg:transfer} with prior $p(\bm{w}) =\mathcal{N}(\bm{\mu}_p,\bm{\Sigma}_p)$ and denote by $(\wh{\bm{\mu}},\wh{\bm{\Sigma}})$ the variational parameter minimizing the objective of Eq.~\eqref{eq:elbo} on a dataset $D$ of $N$ i.i.d. samples distributed according to $\tau$ and $\nu$. Let $\bm{w}^* = \arginf_{\bm{w}}\norm{\wt{B}_{\bm{w}}}_{\nu}^2$ and define $\upsilon(\bm{w}^*) \triangleq \mathbb{E}_{\mathcal{N}(\bm{w}^*,\frac{1}{N}\bm{I})}\left[ \text{v}(\bm{w}) \right]$, with $\text{v}(\bm{w}) \triangleq \mathbb{E}_{\nu}\left[Var_{\mathcal{P}}\left[\wt b(\bm{w})\right]\right]$. Then, there exist constants $c_1,c_2,c_3$ such that, with probability at least $1-\delta$ over the choice of weights $\bm{w} \sim \mathcal{N}(\wh{\bm{\mu}},\wh{\bm{\Sigma}})$ and dataset $D$:
\begin{equation}
\resizebox{0.92\textwidth}{!}{$
\mathbb{E}_{q_{\wh{\bm{\xi}}}}\left[ \norm{\wt B_{\bm{w}}}_{\nu}^2 \right ] \leq\  2\norm{\wt{B}_{\bm{w}^{*}}}_{\nu}^2 + \upsilon(\bm{w}^*) + c_1\sqrt{\frac{\log\frac{2}{\delta}}{N}} + \frac{c_2 + \lambda d \log N + \lambda\norm{\bm{w}^* - \bm{\mu}_p}_{\bm{\Sigma}_p^{-1}}}{N} + \frac{c_3}{N^2}$}.
\end{equation}
\end{restatable}
\begin{proof}
Using Lemma \ref{lemma:general-bound} with variational parameters $\wh{\bm{\xi}} = (\wh{\bm{\mu}},\wh{\bm{\Sigma}})$, we have:
\begin{align}\label{eq:thg-4}
\mathbb{E}_{q_{\wh{\bm{\xi}}}}\left[ \norm{\wt B_{\bm{w}}}_{\nu}^2 \right ] \leq\ & \inf_{\bm{\xi} \in \Xi}\left\{ \mathbb{E}_{q_{\bm{\xi}}}\left[ \norm{\wt B_{\bm{w}}}_{\nu}^2 \right ] + \mathbb{E}_{q_{\bm{\xi}}}\left[ v(\bm{w}) \right] + 2\frac{\lambda}{N} KL({q_{\bm{\xi}}}||p) \right\} + 8\frac{R_{max}^2}{(1-\gamma)^2}\sqrt{\frac{\log\frac{2}{\delta}}{2N}}\notag\\ \leq\ & \mathbb{E}_{\mathcal{N}(\bm{w}^*,c\bm{I})}\left[ \norm{\wt B_{\bm{w}}}_{\nu}^2 \right ] + \mathbb{E}_{\mathcal{N}(\bm{w}^*,c\bm{I})}\left[ v(\bm{w}) \right] + 2\frac{\lambda}{N} KL\left(\mathcal{N}(\bm{w}^*,c\bm{I})\ ||\ p\right)\notag\\ &\ \ + 8\frac{R_{max}^2}{(1-\gamma)^2}\sqrt{\frac{\log\frac{2}{\delta}}{2N}},
\end{align}
where the second inequality is due to the fact that, since Lemma \ref{lemma:general-bound} contains an infimum over the variational parameters, we can upper bound its right-hand side by choosing any specific $\bm{\xi}$ from $\Xi$. Here, we choose $\bm{\mu} = \bm{w}^*$ and $\bm{\Sigma} = c\bm{I}$, for some positive constant $c>0$. Let us now bound these terms separately.

@@

\sum_{j} \chi^{(2)}_{j}KL(\mathcal{N}(\bm{w}^*,c\bm{I})\ ||\ \mathcal{N}(\bm{w}_j, \sigma_p^2\bm{I})) =\ & \frac{1}{2}\sum_{j} \chi^{(2)}_{j}\left( d\log\frac{\sigma^2_p}{c} + d\frac{c}{\sigma^2_p} + \frac{1}{\sigma^2_p}\norm{\bm{w}^* - \bm{w}_j} - d\right)\\ \leq\ & \frac{1}{2}d\log\frac{\sigma^2_p}{c} + \frac{1}{2}d\frac{c}{\sigma^2_p} + \sum_{j} \chi^{(2)}_{j}\frac{1}{2\sigma^2_p}\norm{\bm{w}^* - \bm{w}_j}\\ =\ & \frac{1}{2}d\log\frac{\sigma^2_p}{c} + \frac{1}{2}d\frac{c}{\sigma^2_p} + \varphi\left(\frac{1}{2\sigma^2_p}\norm{\bm{w}^* - \bm{w}_j}\right).

@@

Putting the two terms together:

@@

KL(\mathcal{N}(\bm{w}^*,c\bm{I})\ ||\ p) \leq \log\abs{\mathcal{W}_s} + \frac{1}{2}d\log\frac{\sigma^2_p}{c} + \frac{1}{2}d\frac{c}{\sigma^2_p} + \varphi\left(\frac{1}{2\sigma^2_p}\norm{\bm{w}^* - \bm{w}_j}\right).

@@

Notice that, from now on, one can simply apply the proof of Theorem \ref{th:main-gaussian} with $\sigma_{max}=\sigma_{min}=\sigma^2_p$ and $\frac{1}{2}\norm{\bm{w}^* - \bm{\mu}_p}_{\bm{\Sigma}_p^{-1}}$ replaced by $\varphi\left(\frac{1}{2\sigma^2_p}\norm{\bm{w}^* - \bm{w}_j}\right)$. Thus, by redefining the three constants to $c_1 = \frac{8R_{max}^2}{\sqrt{2}(1-\gamma)^2}$, $c_2 = \bm{w}_{max}^2\bm{\phi}_{max}^2(1 + \gamma)^2 + \lambda d \log\sigma^2_p + 2\lambda\log\abs{\mathcal{W}_s}$, and $c_3 = \frac{1}{2}\gamma^2\kappa^2\bm{\phi}_{max}^4 + \frac{\lambda d}{\sigma^2_p}$, we can write that, with probability at least $1-\delta$:

@@

\mathbb{E}_{q_{\wh{\bm{\xi}}}}\left[ \norm{\wt B_{\bm{w}}}_{\nu}^2 \right ] \leq\  2\norm{\wt{B}_{\bm{w}^{*}}}_{\nu}^2 + \upsilon(\bm{w}^*) + c_1\sqrt{\frac{\log\frac{2}{\delta}}{N}} + \frac{c_2 + \lambda d \log N + 2\lambda\varphi\left(\frac{1}{2\sigma^2_p}\norm{\bm{w}^* - \bm{w}_j}\right)}{N} + \frac{c_3}{N^2}.$}

>>>>

\begin{abstract}
We consider the problem of transferring value functions in reinforcement learning. We propose an approach that uses the given source tasks to learn a prior distribution over optimal value functions and provide an efficient variational approximation of the corresponding posterior in a new target task. We show our approach to be general, in the sense that it can be combined with complex parametric function approximators and distribution models, while providing two practical algorithms based on Gaussians and Gaussian mixtures. We theoretically analyze them by deriving a finite-sample analysis and provide a comprehensive empirical evaluation in four different domains.
\end{abstract}

@@

\begin{restatable}{theorem}{thmg} \label{th:main-mixture}
Fix a target task $\mathcal{M}_{\tau}$. Assume linearly parameterized value functions $Q_{\bm{w}}(s,a)=\bm{w}^T\bm{\phi}(s,a)$ with bounded weights $\norm{\bm{w}}_2\leq w_{max}$ and uniformly bounded features $\norm{\bm{\phi}(s,a)}_2\leq \phi_{\max}$. Consider the mixture version of Algorithm~ \ref{alg:transfer} using $C$ components, source task weights $\mathcal{W}_s$, and bandwidth $\sigma^2_p\ $ for the prior. Denote by $\wh{\bm{\xi}} = (\wh{\bm{\mu}}_1,\dots,\wh{\bm{\mu}}_C,\wh{\bm{\Sigma}}_1,\dots,\wh{\bm{\Sigma}}_C)$ the variational parameters minimizing the objective of Eq. \eqref{eq:elbo} on a dataset $D$ of $N$ i.i.d. samples distributed according to $\tau$ and $\nu$. Let $\bm{w}^* = \arginf_{\bm{w}}||\wt{B}_{\bm{w}}||_{\nu}^2$ and define $\upsilon(\bm{w}^*) \triangleq \mathbb{E}_{\mathcal{N}(\bm{w}^*,\frac{1}{N}\bm{I})}\left[ \text{v}(\bm{w}) \right]$, with $\text{v}(\bm{w}) \triangleq \mathbb{E}_{\nu}\left[Var_{\mathcal{P}_{\tau}}\left[\wt b(\bm{w})\right]\right]$. Then, there exist constants $c_1,c_2,c_3$ such that, with probability at least $1-\delta$ over the choice of the dataset $D$:
\begin{equation}
\resizebox{0.93\textwidth}{!}{$
\mathbb{E}_{q_{\wh{\bm{\xi}}}}\left[ \norm{\wt B_{\bm{w}}}_{\nu}^2 \right ] \leq\  2\norm{\wt{B}_{\bm{w}^{*}}}_{\nu}^2 + \upsilon(\bm{w}^*) + c_1\sqrt{\frac{\log\frac{2}{\delta}}{N}} + \frac{c_2 + \lambda d \log N + 2\lambda\varphi\left(\Delta\right)}{N} + \frac{c_3}{N^2}$},
\end{equation}
where, $\Delta$ is a vector whose $j$-th components is $\Delta = \frac{1}{2\sigma^2_p}\norm{\bm{w}^* - \bm{w}_j}$ and, for a vector $\bm{x} = (x_1,\dots,x_d)$, $\varphi \left( \bm{x}\right) \triangleq \sum_i \frac{e^{-x_i}}{\sum_j e^{-x_j}}x_i$ is the \textit{softmin} function.
\end{restatable}

@@

\begin{restatable}{theorem}{thg} \label{th:main-gaussian}
Fix a target task $\mathcal{M}_{\tau}$. Assume linearly parameterized value functions $Q_{\bm{w}}(s,a)=\bm{w}^T\bm{\phi}(s,a)$ with bounded weights $\norm{\bm{w}}_{2}\leq w_{max}$ and uniformly bounded features $\norm{\bm{\phi}(s,a)}_{2}\leq \phi_{\max}$. Consider the Gaussian version of Algorithm~\ref{alg:transfer} with prior $p(\bm{w}) =\mathcal{N}(\bm{\mu}_p,\bm{\Sigma}_p)$ and denote by $(\wh{\bm{\mu}},\wh{\bm{\Sigma}})$ the variational parameter minimizing the objective of Eq.~\eqref{eq:elbo} on a dataset $D$ of $N$ i.i.d. samples distributed according to $\tau$ and $\nu$. Let $\bm{w}^* = \arginf_{\bm{w}}\norm{\wt{B}_{\bm{w}}}_{\nu}^2$ and define $\upsilon(\bm{w}^*) \triangleq \mathbb{E}_{\mathcal{N}(\bm{w}^*,\frac{1}{N}\bm{I})}\left[ \text{v}(\bm{w}) \right]$, with $\text{v}(\bm{w}) \triangleq \mathbb{E}_{\nu}\left[Var_{\mathcal{P}}\left[\wt b(\bm{w})\right]\right]$. Then, there exist constants $c_1,c_2,c_3$ such that, with probability at least $1-\delta$ over the choice of the dataset $D$:
\begin{equation}
\resizebox{0.92\textwidth}{!}{$
\mathbb{E}_{q_{\wh{\bm{\xi}}}}\left[ \norm{\wt B_{\bm{w}}}_{\nu}^2 \right ] \leq\  2\norm{\wt{B}_{\bm{w}^{*}}}_{\nu}^2 + \upsilon(\bm{w}^*) + c_1\sqrt{\frac{\log\frac{2}{\delta}}{N}} + \frac{c_2 + \lambda d \log N + \lambda\norm{\bm{w}^* - \bm{\mu}_p}_{\bm{\Sigma}_p^{-1}}}{N} + \frac{c_3}{N^2}$}.
\end{equation}
\end{restatable}
\begin{proof}
Using Lemma \ref{lemma:general-bound} with variational parameters $\wh{\bm{\xi}} = (\wh{\bm{\mu}},\wh{\bm{\Sigma}})$, we have:
\begin{align}\label{eq:thg-4}
\mathbb{E}_{q_{\wh{\bm{\xi}}}}\left[ \norm{\wt B_{\bm{w}}}_{\nu}^2 \right ] \leq\ & \inf_{\bm{\xi} \in \Xi}\left\{ \mathbb{E}_{q_{\bm{\xi}}}\left[ \norm{\wt B_{\bm{w}}}_{\nu}^2 \right ] + \mathbb{E}_{q_{\bm{\xi}}}\left[ v(\bm{w}) \right] + 2\frac{\lambda}{N} KL({q_{\bm{\xi}}}||p) \right\} + 8\frac{R_{max}^2}{(1-\gamma)^2}\sqrt{\frac{\log\frac{2}{\delta}}{2N}}\notag\\ \leq\ & \mathbb{E}_{\mathcal{N}(\bm{w}^*,c\bm{I})}\left[ \norm{\wt B_{\bm{w}}}_{\nu}^2 \right ] + \mathbb{E}_{\mathcal{N}(\bm{w}^*,c\bm{I})}\left[ v(\bm{w}) \right] + 2\frac{\lambda}{N} KL\left(\mathcal{N}(\bm{w}^*,c\bm{I})\ ||\ p\right)\notag\\ &\ \ + 8\frac{R_{max}^2}{(1-\gamma)^2}\sqrt{\frac{\log\frac{2}{\delta}}{2N}},
\end{align}
where the second inequality is due to the fact that, since Lemma \ref{lemma:general-bound} contains an infimum over the variational parameters, we can upper bound its right-hand side by choosing any specific $\bm{\xi}$ from $\Xi$. Here, we choose $\bm{\mu} = \bm{w}^*$ and $\bm{\Sigma} = c\bm{I}$, for some positive constant $c>0$. Let us now bound these terms separately.

@@

\sum_{j} \chi^{(2)}_{j}KL(\mathcal{N}(\bm{w}^*,c\bm{I})\ ||\ \mathcal{N}(\bm{w}_j, \sigma_p^2\bm{I})) =\ & \frac{1}{2}\sum_{j} \chi^{(2)}_{j}\left( d\log\frac{\sigma^2_p}{c} + d\frac{c}{\sigma^2_p} + \frac{1}{\sigma^2_p}\norm{\bm{w}^* - \bm{w}_j} - d\right)\\ \leq\ & \frac{1}{2}d\log\frac{\sigma^2_p}{c} + \frac{1}{2}d\frac{c}{\sigma^2_p} + \sum_{j} \chi^{(2)}_{j}\frac{1}{2\sigma^2_p}\norm{\bm{w}^* - \bm{w}_j}\\ =\ & \frac{1}{2}d\log\frac{\sigma^2_p}{c} + \frac{1}{2}d\frac{c}{\sigma^2_p} + \varphi\left(\Delta\right).

@@

where we defined the vector $\Delta$ whose components are $\Delta_j = \frac{1}{2\sigma^2_p}\norm{\bm{w}^* - \bm{w}_j}$. Putting the two terms together:

@@

KL(\mathcal{N}(\bm{w}^*,c\bm{I})\ ||\ p) \leq \log\abs{\mathcal{W}_s} + \frac{1}{2}d\log\frac{\sigma^2_p}{c} + \frac{1}{2}d\frac{c}{\sigma^2_p} + \varphi\left(\Delta\right).

@@

Notice that, from now on, one can simply apply the proof of Theorem \ref{th:main-gaussian} with $\sigma_{max}=\sigma_{min}=\sigma^2_p$ and $\frac{1}{2}\norm{\bm{w}^* - \bm{\mu}_p}_{\bm{\Sigma}_p^{-1}}$ replaced by $\varphi\left(\Delta\right)$. Thus, by redefining the three constants to $c_1 = \frac{8R_{max}^2}{\sqrt{2}(1-\gamma)^2}$, $c_2 = \bm{w}_{max}^2\bm{\phi}_{max}^2(1 + \gamma)^2 + \lambda d \log\sigma^2_p + 2\lambda\log\abs{\mathcal{W}_s}$, and $c_3 = \frac{1}{2}\gamma^2\kappa^2\bm{\phi}_{max}^4 + \frac{\lambda d}{\sigma^2_p}$, we can write that, with probability at least $1-\delta$:

@@

\mathbb{E}_{q_{\wh{\bm{\xi}}}}\left[ \norm{\wt B_{\bm{w}}}_{\nu}^2 \right ] \leq\  2\norm{\wt{B}_{\bm{w}^{*}}}_{\nu}^2 + \upsilon(\bm{w}^*) + c_1\sqrt{\frac{\log\frac{2}{\delta}}{N}} + \frac{c_2 + \lambda d \log N + 2\lambda\varphi\left(\Delta\right)}{N} + \frac{c_3}{N^2}.$}
